Changes in accessibility to foodbanks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario

### 2021-04-22 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
Freshly knitted after merging changes

journal: "An awesome journal"
date: "`r Sys.Date()`"
bibliography: references.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

@@

**The locations of foodbanks were obtained from public records and geocoded. Three urban sites not in the program were also identified and geocoded for comparison purposes.**

@@

### Open Hamilton **???**

We used the open data portal of the City of Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained boundaries for the city's various regions (the definition of urban, suburban, and rural regions follows the classification of development applications).

@@

Once we obtained travel time tables with population (number of households) and proportion of trips by mode, we calculated the expected travel time $ett$ from each parcel $i$ to a foodbank $j$ as follows:
$$
ett_i = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
$$
\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the travel time from parcel $i$ to the foodbank. In other words, the expected travel time is the weighted sum of travel times to the foodbank, with the weights given by the expected modal split in the TAZ.

>>>>

journal: "An awesome journal"
date: "`r Sys.Date()`"
bibliography: references.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
    - \usepackage[table]{xcolor}
---

@@

The locations of foodbanks were obtained from public records and geocoded. Three urban sites not in the program were also identified and geocoded for comparison purposes.

@@

@@

Once we obtained travel time tables with population (number of households) and proportion of trips by mode, we calculated the expected travel time $ett$ from each parcel $i$ to a foodbank $j$ as follows:
$$
ett_i = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
$$

\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the travel time from parcel $i$ to the foodbank. In other words, the expected travel time is the weighted sum of travel times to the foodbank, with the weights given by the expected modal split in the TAZ.

### 2021-04-22 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
Freshly knitted after merging changes

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

It is known that the urban food environment, within which people make
their daily food choices, is essential in influencing eating behaviours
and health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Although
previous work has explored differences in \emph{accessing} food banks,
such as how some households utilize food banks over short periods of
time while others regularly utilize food banks as longer-term resource
(e.g., Enns et al., 2020), we are not aware of any research that has
focused on the geographic component of accessibility. \textbf{Check
Tarasuk et al. (2019)}

@@

\hypertarget{literature-review}{%
\section{Literature Review}\label{literature-review}}

\hypertarget{food-insecurity}{%
\subsection{Food Insecurity}\label{food-insecurity}}

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to inadequate nutrient intake
(Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt, 2009). This nutrient
deficiency has been causing major health concerns in Canadians, and
particularly those who are at a socioeconomic disadvantage (Bazerghi et
al., 2016). Previous studies have aimed to look at the relationship
between the built food environment and socio-demographic characteristics
with qualitative and quantitative, or a mixed -method approach.
Quantitatively, official governmental surveys have been able to assist
with data, such as the Household Food Security Survey Module (HFSSM),
the Canadian Community Health Surveys (CCHS), the Longitudinal and
International Study of Adults (LISA), and official classifications
determined by Health Canada in relation to sociodemographic variables
(Gundersen et al., 2018; Kirkpatrick and Tarasuk, 2008; Tarasuk and
Vogt, 2009). Studies have also aimed to assess food availability of
healthy foods (e.g., fruit and vegetables) at supermarkets in relation
to sociodemographic characteristics and geographic accessibility (Latham
and Moffat, 2007). In terms of findings, studies have generated
inconsistent relationships between their evaluated availability of food
and sociodemographic characteristics that go in hand.

\hypertarget{foodbanks}{%
\subsection{Foodbanks}\label{foodbanks}}

Food banks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). The scope and objectives of
foodbanks can vary by region and by country, and these organizations can
include not only prepared meals and aliments, but also shared spaces to
connect in community gardens and community kitchens (Wakefield et al.,
2013). Although in their origin foodbanks were meant to be provide a
temporary solution to accommodate those in hunger due to job
retrenchments and economic downfalls since the 1980s, over time they
evolved into a community practice to secure emergency food supplies for
those in need (Loopstra and Tarasuk, 2012; Wakefield et al., 2013). In
Canada, the number of foodbanks has steadily increased in the past few
decades (Wakefield et al., 2013). Previous studies have questioned
whether food banks are able to offer a rounded nutritious supply of
food, and if food banks are a sustainable on-going practice for those in
need (Bazerghi et al., 2016; Riches, 2002). An inescapable reality, at
the moment, is that despite their imperfections they provide urgently
needed support for members of the population for whom foodbanks are
frequently their primary location to get food (Bazerghi et al., 2016).
\textbf{However, surveys revealed that only 20 to 30 percent of those
experiencing food insecurity were found to frequent food banks in Canada
(Tarasuk et al., 2014).} \textbf{Something about geographical access}

\hypertarget{food-insecurity-in-canada-during-the-covid-19}{%
\subsection{Food Insecurity in Canada during the
COVID-19}\label{food-insecurity-in-canada-during-the-covid-19}}

Currently with the COVID-19 pandemic, disrupted economies, rising
unemployment rates, and changes in poverty levels have disrupted the
food environment and led to higher rates of food insecurity (Niles et
al., 2020). In 2012, 12.4\% of Canadians households and 11.8\% Ontarian
households experienced some degree of food insecurity (Gundersen et al.,
2018; Tarasuk et al., 2020). Pandemic-related food security studies in
the US have found a substantial increase in households experiencing food
insecurity for the first time, and also in households experiencing more
severe food insecurity than before (Niles et al., 2020; Wolfson and
Leung, 2020). Most recently in May 2020, Canada recorded 14.7\% of its
population living in food insecurity in the past 30 days (Statistics
Canada, 2020a). Food insecurity is a highly concerning public health
issue due to its health impacts. Recent research suggests that food
insecurity in adults can lead to experiencing more stressful events
(El-hajj and Benhin 2021). Increases in food insecurity rates due to the
pandemic, signal a change in the food environment with potential damages
to population health during the course of the pandemic and beyond (Niles
et al., 2020).

@@

\textbf{The locations of foodbanks were obtained from public records and
geocoded. Three urban sites not in the program were also identified and
geocoded for comparison purposes.}

@@

\hypertarget{open-hamilton}{%
\subsubsection{\texorpdfstring{Open Hamilton
\textbf{???}}{Open Hamilton ???}}\label{open-hamilton}}

We used the open data portal of the City of
Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained
boundaries for the city's various regions (the definition of urban,
suburban, and rural regions follows the classification of development
applications).

@@

Once we obtained travel time tables with population (number of
households) and proportion of trips by mode, we calculated the expected
travel time \(ett\) from each parcel \(i\) to a foodbank \(j\) as
follows: \[
ett_i = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
\] \noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the travel time from
parcel \(i\) to the foodbank. In other words, the expected travel time
is the weighted sum of travel times to the foodbank, with the weights
given by the expected modal split in the TAZ.

@@

\leavevmode\hypertarget{ref-latham2007determinants}{}%
Latham, J., Moffat, T., 2007. Determinants of variation in food cost and
availability in two socioeconomically contrasting neighbourhoods of
hamilton, ontario, canada. Health \& Place 13, 273--287.
doi:\url{https://doi.org/10.1016/j.healthplace.2006.01.006}

@@

>>>>

% Pandoc header
\usepackage[table]{xcolor}

@@

It is known that the urban food environment, within which people make
their daily food choices, is essential in influencing eating behaviours
and health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Although
previous work has explored differences in \emph{accessing} food banks,
such as how some households utilize food banks over short periods of
time while others regularly utilize food banks as longer-term resource
(e.g., Enns et al., 2020), we are not aware of any research that has
focused on the geographic component of accessibility.

@@

\hypertarget{food-insecurity-and-food-bank-use-in-canada}{%
\section{Food Insecurity and Food Bank Use in
Canada}\label{food-insecurity-and-food-bank-use-in-canada}}

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to inadequate nutrient intake (Enns
et al., 2020; Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt, 2009).
This nutrient deficiency is associated with major health concerns in
Canadians, and particularly those who are at a socioeconomic
disadvantage (Bazerghi et al., 2016). Data on food insecurity in Canada
is collected in several ways. Quantitatively, official government
surveys such as the Household Food Security Survey Module (HFSSM), the
Canadian Community Health Surveys (CCHS), the International Study of
Adults (LISA), and official classifications determined by Health Canada
in relation to sociodemographic variables offer some insight into food
insecurity (Gundersen et al., 2018; Kirkpatrick and Tarasuk, 2008;
Tarasuk and Vogt, 2009). Nationally, it has been found that food
insecurity impacts approximately 12.3\% of Canadian households (Tarasuk
et al., 2014)

Food banks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). The scope and objectives of
foodbanks can vary by region and by country, and these organizations can
include not only prepared meals and aliments, but also shared spaces to
connect in community gardens and community kitchens (Wakefield et al.,
2013). Although in their origin foodbanks were meant to be provide a
temporary solution to accommodate those in hunger due to job
retrenchments and economic downfalls since the 1980s, over time they
evolved into a community practice to secure emergency food supplies for
those in need (Loopstra and Tarasuk, 2012; Wakefield et al., 2013).

In Canada, the number of food banks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of food banks
and their use comes from the non-profit association Food Banks Canada,
which conducts an annual assessment through its affiliated members. Food
Banks Canada's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member food banks across the
country found that Canadians visited food banks 1.1 million times in
March of 2018. Of those accessing food banks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to Food Banks
Canada's 2018 data, single-adult households represent 45\% of those
utilizing food banks despite making up 28\% of Canada's population, 19\%
are single-parent households (compared to 10\% nationally), and 35\% of
those accessing food bank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing food banks list social or
disability assistance as their primary source of income. Moreover,
children aged 0-18 are over-represented as food bank users, making up
35\% of those accessing food bank services despite representing 20\% of
the country's population. Likewise, using data from the 2011-2012 CCHS,
Tarasuk et al. (2019) found higher odds of food insecurity amongst
households relying on social assistance, those without a university
degree or with children under the age of 18, and individuals that lived
alone, renters, and those identifying as Aboriginal. While surveys
revealed that only 20 to 30 percent of those experiencing food
insecurity were found to frequent food banks in Canada (Tarasuk et al.,
2014), research from Ottawa (Enns et al., 2020) and Vancouver (Black and
Seto, 2020) suggests that long-term users tend to be older, have health
or mobility challenges, live in large households, and less likely to
have employment income.

Currently with the COVID-19 pandemic, disrupted economies, rising
unemployment rates, and changes in poverty levels have disrupted the
food environment and led to higher rates of food insecurity (Niles et
al., 2020). Pandemic-related food security studies in the US have found
a substantial increase in households experiencing food insecurity for
the first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from Food Banks Canada showed that 52\% of member food banks reported an
increase in usage in March of 2020 when initial lockdown restrictions
were put in place across much of Canada. The pandemic also created
significant staffiing issues with 42\% of food banks reporting a
reduction in volunteers. However, 53\% of food banks reported a decrease
in use into the summer of 2020 which food banks attributed to emergency
financial support programs from the federal government were.
Nevertheless, some of these benefit programs were temporary, suggesting
many households may again turn to food bank services to meet their
needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing food banks, the locational or
transportation accessibility aspect of food bank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socioeconomic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For food banks specifically, previous qualitative
research in Ontario has noted difficulties associated with
transportation for trips to the food bank (Smith-Carrier et al., 2017).
For example, Smith et al. (2017) note that participants access food
banks using a variety of modes of transportation with respondent stating
``I wish it {[}the food bank{]} was a little more centrally located.
Because if I didn't have a bike I'd have to walk it all the way out
there and back. I wonder about people who don't.'' From this, the
authors conclude that ``transportation can be challenging, particularly
if the food bank is situated in a remote location.'' To offer greater
insight into the role of transportation and location in food bank
accessibility, this research examines how geographic access to food
banks and food services changed in Hamilton before and during the
COVID-19 pandemic.

@@

The locations of foodbanks were obtained from public records and
geocoded. Three urban sites not in the program were also identified and
geocoded for comparison purposes.

@@

@@

Once we obtained travel time tables with population (number of
households) and proportion of trips by mode, we calculated the expected
travel time \(ett\) from each parcel \(i\) to a foodbank \(j\) as
follows: \[
ett_i = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
\]

\noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the travel time from
parcel \(i\) to the foodbank. In other words, the expected travel time
is the weighted sum of travel times to the foodbank, with the weights
given by the expected modal split in the TAZ.

@@

@@

\leavevmode\hypertarget{ref-smith2017food}{}%
Smith-Carrier, T., Ross, K., Kirkham, J., Decker Pierce, B., 2017.
{`Food is a right\ldots{} nobody should be starving on our streets'}:
Perceptions of food bank usage in a mid-sized city in ontario, canada.
Journal of Human Rights Practice 9, 29--49.
